Tropical mathematics for the Lambda-calculus

### 2023-07-25 CSL24/main.tex
Commit
Commit giornaliero
Committino notturno

\title{Linearity and tropical mathematics in the Lambda-calculus} %TODO Please add

@@

\section{Introduction}
\input{section1}

%\section{The Two Quantitative Approaches to the $\lambda$-calculus}\label{section2}
\input{section2}

@@

{\color{red}2 pagine in totale}

\subsection{Quantitaive analysis: probabilistic, differential privacy}\label{sec:app}
%\input{section5}
%\input{section5C}

>>>>

\title{Tropical mathematics for the Lambda-calculus} %TODO Please add

@@

\section{Two kinds of quantitative approaches}
\input{section1}

%\section{The Two Quantitative Approaches to the $\lambda$-calculus}\label{section2}
%\input{section2}

@@

{\color{red}3 pagine in totale}

\subsection{Quantitaive analysis%: probabilistic, differential privacy
}\label{sec:app}
\input{section5}
%\input{section5C}

### 2023-07-25 CSL24/section4.tex
Committino notturno

 In a similar way, tLs
$f:\Lawv^X\to\Lawv^Y$ 
 with \emph{finite} supports $\C F_b=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$, and which have thus shape $f(x)_b=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, are generalisation of usual tropical polynomials to the case of possibly infinitely many variables.

% 
% , i.e.~for a \emph{finite} set $\C F\subseteq \, !X$.
%%Remark that we also find usual \emph{tropical polynomials} of tropical geometry as a particular case: they 
%correspond to the tLs for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}. 
%Actually, for us \emph{tropical polynomial} will mean a function 
%\end{remark}

@@

\begin{corollary}
Let $\model A$ be a finite set.
\begin{enumerate}
\item $\model{\Gamma \vdash_{\BSTLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{tropical polynomial}, thus (as $\model A$ is finite), a \emph{Lipschitz} function.
\item $\model{\Gamma \vdash_{\STLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{locally} Lipschitz map.
\item $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!$ of \emph{tropical polynomials}, thus (as $\model A$ is finite), \emph{Lipschitz} functions.
\end{enumerate}
\end{corollary} 
\begin{proof}
1). As we observed, the interpretation $\model A$ of a bounded type is a finite set, hence $\model M$ must be a finite matrix  and  its corresponding linear map $\Lawv^{\model A} \to \Lawv^{\model B}$ is thus a tropical polynomial, so we apply Corollary~\ref{prop:polylip}.
2). It follows directly from Theorem~\ref{thmTLSlocLip}.
3). It follows from the easily checked fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
\end{proof}

>>>>

 In a similar way, tLs
$f:\Lawv^X\to\Lawv^Y$ 
 with \emph{finite} supports $\C F_b=\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$, and which have thus shape $f(x)_b=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, are generalisation of usual tropical polynomials to the case of possibly infinitely many variables.
For instance, for a $t\in\HOM{\LREL}{!_n X}{Y}$ (this is in particular the case for the interpretations of $\BSTLC$-terms), we will consider the (generalised) tropical polynomial $t^!:\Lawv^X\to\Lawv^Y$ defined by $t^!(x)_b:=\inf_{\mu\in !_nX}\{\mu x+t_{\mu,b}\}$.
% 
% , i.e.~for a \emph{finite} set $\C F\subseteq \, !X$.
%%Remark that we also find usual \emph{tropical polynomials} of tropical geometry as a particular case: they 
%correspond to the tLs for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}. 
%Actually, for us \emph{tropical polynomial} will mean a function 
%\end{remark}

@@

\begin{corollary}
Let $\model A$ be a finite set.
\begin{enumerate}
\item $\model{\Gamma \vdash_{\BSTLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{tropical polynomial}, thus (as $\model A$ is finite), a \emph{Lipschitz} function.
\item $\model{\Gamma \vdash_{\STLC} M:A}^!:\Lawv^{\model\Gamma} \to \Lawv^{\model A}$ is a \emph{locally} Lipschitz map.
\item $\Te{M}$ decomposes $\model{\Gamma \vdash_{\STLC} M:A}^!$ as an $\inf_{t\in\Te{M}}\model{\Gamma\vdash_{\STDLC} t:A}^!$ of \emph{tropical polynomials}, thus (as $\model A$ is finite), \emph{Lipschitz} functions.
\end{enumerate}
\end{corollary} 
\begin{proof}
1). We already observed that the interpretation of a $\BSTLC$-type $B$ is alway a finite set, so thus is $!_n \model B$.
So the $\inf$ defining $(\_)^!$ is actually a $\min$.
So the interpretation of a bounded term is a tropical polynomial, so we apply Corollary~\ref{prop:polylip} to each coordinate of the image, and since $\model A$ is finite (by taking the maximum Lipschitz constant among the finite number of them), we obtain the thesis.
2). It follows immediately from Theorem~\ref{thmTLSlocLip} and the fact that $\model A$ is finite.
3). It follows from \autoref{cor:T(M)=M} plus the easily checked fact that, for $(f_n)_{n\in\N}\subseteq\Lawv^{!X\times Y}$, we have $\left(\inf_{n\in\N} f_n\right)^!:\Lawv^X\to\Lawv^Y$, with $\left(\inf_{n\in\N} f_n\right)^!=\inf_{n\in\N} f_n^!$.
\end{proof}

### 2023-07-25 CSL24/section5.tex
Committino notturno

\begin{remark}
 Now the matrix interpreting a term need no longer be boolean.
 For instance, $\model{\vdash_{\STLC_\oplus} \true:\bool}_i$ is $p$ if $i=1$ and it is $1-p$ if $i=0$.
 In general, one can show that $\model{\Gamma \vdash_{\STLC_\oplus} M:A}\in[0,1]^{!\model \Gamma \times \model A}$.
 Coefficients $> 1$ may be considered when dealing with other kinds of effects.
\end{remark}

>>>>

\begin{remark}
 Now the matrix interpreting a term need no longer be boolean.
 For instance, $\model{\vdash_{\STLC_\oplus} \true\oplus_p\false :\bool}_i$ is $p$ if $i=1$ and it is $1-p$ if $i=0$.
 %In general, one can show that $\model{\Gamma \vdash_{\STLC_\oplus} M:A}\in[0,1]\cup\{\infty\}^{!\model \Gamma \times \model A}$. Coefficients $> 1$ may be considered when dealing with other kinds of effects.
\end{remark}
